Assessing Gender Bias in Machine Translation -- A Case Study with Google Translate

### 2017-12-26 paper/main.tex
Fixed problem with JAIR format

\documentclass{article}

\usepackage[utf8]{inputenc}
%\usepackage{fontspec}
\usepackage{xeCJK}
%\usepackage{armtex}

@@

\title{Is Google Translate Sexist? -- Assessing Gender Bias in Machine Translation}
\author{Marcelo Prates \and Luis Lamb}

@@

\begin{abstract}
In recent times there has been a growing concern both inside and outside academia about the phenomenon of machine bias, where trained statistical models unbeknownst to their creators become vessels of their own prejudices. A large number of AI tools have recently been suggested to be harmfully biased towards some minority, with reports of racist criminal behavior predictors, Apple's Iphone X failing to differentiate between two Asian people and the infamous case of Google photos' classifying black people as gorillas. Although a systematic study of such biases can be difficult, we believe that automated translation tools can be exploited through gender neutral languages to yield a window into the phenomenon of \emph{gender} bias in AI. In this paper, we construct sentences with gender neutral pronouns in different languages which support them and translate these sentences into English using the Google Translate API. The resulting English sentences are constructed either with male, female or neutral pronouns, which allows us to collect statistics about the frequency of male defaults and evaluate which factors can influence the gender Google Translate assumes for each sentence. We show that not only there is a tendency towards translating pronouns to \say{he} but also that this tendency is exaggerated for sentences about male dominated fields such as computer science. At the other hand, sentences about artistic fields are less likely to yield male defaults. This possibly reflects the implicit assumptions our society has about stereotypical gender roles, which suggests that without proper care statistical machine translation can become hostage to the prejudices of those people behind its training dataset -- ourselves.
\end{abstract}

@@

Although the idea of automated translation can in principle be traced back to as long as the 17th century with René Descartes proposal of an \say{universal language} \cite{dascal1982universal}, machine translation has only existed as a technological field since the 1950s, with a pioneering memorandum by Warren Weaver \cite{locke1955machine} discussing the possibility of employing digital computers to perform automated translation. The now famous Georgetown-IBM experiment followed not long after, providing the first experimental demonstration of the prospects of automating translation by the means of successfully converting more than sixty Russian sentences into English \cite{gordin2015scientific}. Early systems improved upon the results of the Georgetown-IBM experiment by exploiting Chomsky's theory of generative linguistics, and the field experienced a sense of optimism about the prospects of fully automating natural language translation. As is customary with artificial intelligence, the initial optimistic stage was followed by an extended period of strong disillusionment with the field, of which the catalyst was the influential ALPAC report \cite{hutchins1986machine}. Research was almost completely abandoned in the United States, making a shy re-entrance in the 1970s before the 1980s surge in statistical methods for machine translation \cite{koehn2009statistical}. Statistical and example-based machine translation have been on the rise ever since \cite{carl2003recent}, with highly successful applications such as Google Translate (recently ported to a neural translation technology \cite{wu2016google,gtranslateneuralnet2017}) amounting to over $200$ million users daily \cite{gtranslateneuralnet2017,gtranslatezeroshot2017}.

In spite of the recent commercial success of automated translation tools (or perhaps stemming directly from it), machine translation has amounted a significant deal of criticism. Noted philosopher and founding father of generative linguistics Noam Chomsky has argued that the achievements of machine translation, while successes in a particular sense, are \emph{not successes in the sense that science has ever been interested in}: they merely provide effective ways, according to Chomsky, of approximating unanalyzed data. Chomsky argues that the faith of the MT community in statistical methods is absurd by analogy with a standard scientific field such as physics:

\begin{quotation}
I mean actually you could do physics this way, instead of studying things like balls rolling down frictionless planes, which can't happen in nature, if you took a ton of video tapes of what's happening outside my office window, let's say, you know, leaves flying and various things, and you did an extensive analysis of them, you would get some kind of prediction of what's likely to happen next, certainly way better than anybody in the physics department could do. Well that's a notion of success which is I think novel, I don't know of anything like it in the history of science.
\end{quotation}

Leading AI researcher and Google's Director of Research Peter Norvig responds to these arguments by suggesting that even standard physical theories such as the Newtonian model of gravitation are, in a sense, \emph{trained}:

\begin{quotation}
As another example, consider the Newtonian model of gravitational attraction, which says that the force between two objects of mass $m_1$ and $m_2$ a distance $r$ apart is given by

\begin{equation*}
F = G m_1 m_2 / r^2
\end{equation*}

where $G$ is the universal gravitational constant. This is a trained model because the gravitational constant G is determined by statistical inference over the results of a series of experiments that contain stochastic experimental error. It is also a deterministic (non-probabilistic) model because it states an exact functional relationship. I believe that Chomsky has no objection to this kind of statistical model. Rather, he seems to reserve his criticism for statistical models like Shannon's that have quadrillions of parameters, not just one or two.
\end{quotation}

Chomsky and Norvig's debate \cite{norvig2017chomsky} is a microcosmos of the two leading standpoints about the future of science in the face of increasingly sophisticated statistical models. Are we, as Chomsky seems to argue, jeopardizing science by relying on statistical tools to perform predictions instead of perfecting traditional science models, or are these tools, as Norvig argues, components of the scientific standard since its conception? Currently there are no satisfactory resolutions to this conundrum, but perhaps statistical models pose an even greater and more urgent threat to our society. On a 2014 article, Londa Schiebinger suggested that scientific research fails to take gender issues into account, arguing that the phenomenon of male defaults on new technologies such as Google Translate provides a window into this asymmetry \cite{schiebinger2014scientific}. Since then, recent worrisome results in machine learning have somewhat supported Schiebinger's view, and perhaps even partially confirmed some of Chomsky's fears. Not only Google photos' statistical image labeling algorithm has been found to classify dark-skinned people as gorillas \cite{garcia2016racist} and purportedly intelligent programs have been suggested to be negatively biased against black prisoners when predicting criminal behavior \cite{angwin2016machine} but the machine learning revolution has also indirectly revived heated debates about the controversial field of physiognomy, with proposals of AI systems capable of identifying the sexual orientation of an individual through its facial characteristics \cite{wang2017deep}. Similar concerns are growing at an unprecedented rate in the media, with reports of Apple's Iphone X face unlock feature failing to differentiate between two different Asian people \cite{womanunlockphone2017} and automatic soap dispensers which reportedly do not recognize black hands \cite{racistsoapdispenser2017}. \emph{Machine bias}, the phenomenon by which trained statistical models unbeknownst to their creators become vessels of their own prejudices, is growing into a pressing concern for the modern times, and is an invitation for us to ask ourselves whether there are limits to our dependence on these techniques -- and more importantly, whether some of these limits have already been traversed.

With this in mind, we propose a quantitative analysis of the phenomenon of gender bias in machine translation. We believe this can be done by simply exploiting Google Translate to map sentences from a gender neutral language to English. As Figure \ref{fig:screenshot-gtranslate-hungarian} exemplifies, this approach produces results consistent with the hypothesis that senteces about stereotypical gender roles are translated accordingly with high probability: \emph{nurse} and \emph{baker} are translated with female pronouns while \emph{engineer} and \emph{CEO} are translated with male ones.

@@

If conclusive, in our view such an analysis would suggest that automatic translation tools should be designed in such a way as to avoid becoming hostages to their own underlying statistical models. Because these tools are trained with real world data they implicitly absorb the biases and stereotypes of our society, which must then be painstakingly removed. This poses an ethical rather than a technical limitation to statistical machine translation and big data statistical models in general.

\section{Methods}

We believe that the phenomenon of gender bias in machine translation can be assessed by mapping sentences constructed in gender neutral languages to English by the means of an automated translation tool. Specifically, we can translate sentences such as the Hungarian \say{ő ápolónő}, where \say{ápolónő} translates to \say{nurse} and \say{ő} is a gender-neutral pronoun meaning either he, she or it, to English, yielding in this example the result \say{she's a nurse} on Google Translate. The same basic template can be ported to all other gender neutral languages, as Table \ref{tab:templates} shows. Given the success of Google Translate, which amounts to 200 million users daily, we have decided to exploit its API to obtain the desired thermometer of gender bias. Also, in order to solidify our results, we have decided to work with as many gender neutral languages as possible, obtaining a list of these from the Wikipedia article on the issue (\url{https://en.wikipedia.org/wiki/Gender_neutrality_in_genderless_languages}). Table \ref{tab:gender-neutral-languages} compiles all languages from said article, with additional columns informing whether they 1) exhibit a pronominal gender system and 2) are supported by Google Translate. Because pronominal gender systems defy the purposes of our technique, such languages have been discarded. Following difficulties with Bengali, Nepali and Korean we have decided not to work with these languages also.

@@

\begin{table}[H]
	\centering
	\begin{tabular}{|c|m{2cm}|m{2cm}|c|c|}
	\hline
	Language Family & Language & Pronominal Gender System & Supported & Tested \\ \hline \hline
	\multirow{2}{*}{Austronesian} 	& Malay 				& \xmark 			& \checkmark	& \checkmark	\\
									& Tagalog 				& \xmark 			& \xmark		& \xmark		\\ \hline
	\multirow{3}{*}{Finno-Ugric} 	& Estonian 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Finnish 				& \xmark 			& \checkmark	& \checkmark	\\
									& Hungarian 			& \xmark 			& \checkmark 	& \checkmark	\\ \hline
	\multirow{4}{*}{Indo-European} 	& Armenian 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Bengali 				& \xmark 			& \checkmark 	& \xmark		\\
									& English 				& \checkmark 		& \checkmark 	& \xmark 		\\
									& Persian 				& \checkmark 		& \checkmark 	& \xmark 		\\ \hline
	\multirow{3}{*}{Indo-Aryan} 	& Maithili 				& \xmark 			& \xmark 		& \xmark 		\\
									& Nepali 				& \xmark 			& \checkmark	& \xmark		\\
									& Oriya 				& \xmark 			& \xmark 		& \xmark 		\\ \hline
	\multirow{10}{*}{} 				& Japanese 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Korean 				& O 				& \checkmark 	& \xmark		\\
									& Turkish 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Yoruba 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Basque 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Swahili 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Chinese 				& O 				& \checkmark 	& \checkmark	\\
									& Cantonese 			& \xmark 			& \xmark		& \xmark		\\
									& Pipil 				& \xmark 			& \xmark		& \xmark		\\
									& Quechuan 				& \xmark 			& \xmark		& \xmark		\\ \hline
	\multirow{4}{*}{Constructed} 	& Esperanto 			& \checkmark 		& \checkmark 	& \xmark		\\
									& Ido 					& O 				& \xmark 		& \xmark 		\\
									\cline{2-2}
									& Lingua Franca Nova 	& \xmark 			& \xmark 		& \xmark 		\\
									\cline{2-2}
									& Interlingua 			& \xmark 			& \xmark 		& \xmark 		\\ \hline
	\end{tabular}
	\caption{Selected gender neutral languages obtained from the Wikipedia article \url{https://en.wikipedia.org/wiki/Gender_neutrality_in_genderless_languages}. Languages are grouped according to language families and classified according to whether they exhibit pronominal gender system (\checkmark: yes, \xmark:~no, O: it is optional). For the purposes of this work, we have decided to work only with languages lacking such a system, and as such Persian and Esperanto have been discarded. Languages lacking support from Google Translate have been discarded. Following difficulties with Bengali, Nepali and Korean, these languages have also been discarded.}
	\label{tab:gender-neutral-languages}
\end{table}

@@

Figures \ref{fig:histogram-female}, \ref{fig:histogram-female} and \ref{fig:histogram-neutral} shed further light on the asymmetry between the distribution male and female pronouns. While the number of occurences of male pronouns tends towards a normal distribution, the figure is changes drastically for the opposite gender, whose pronoun distribution is apparently governed by an inverse correlation. The behavior repeats itself for the gender neutral pronouns in Figure \ref{fig:histogram-neutral}.

@@

\section{Discussion}

In this paper, we have provided evidence that statistical translation tools such as Google Translate exhibit implicit gender biases and a tendency towards male defaults, which possibly stem from the real world data used to train them. As a result, we suggest that such tools can be probed to yield insights about stereotypical gender roles in our society. By translating sentences from gender neutral languages such as Hungarian and Chinese into English, we were able to collect statistics about the asymmetry between female and male pronominal genders in the translation process. Because Google Translate typically uses English as a \emph{lingua franca} to translate between other languages, our findings probably generalize to most translations from gender neutral idioms, although we have not tested this hypothesis. Our results further show that although male pronouns are typical, the proportion of female pronouns can vary significantly according either to the adjectives used (\say{shy}/\say{happy} vs \say{brave}/\say{Guilty}) or the category of professional occupations (artistic jobs are far more likely to be translated with female pronouns than computer-related ones). We have also shown that different languages are differently biased towards male defaults, with Hungarian exhibiting a better balance between male and female pronouns than, say, Chinese. Some languages such as Yoruba and Basque have been found to translate sentences with gender neutral pronouns very often, although this is the exception rather than the rule.

We think this work can shed further light on some of the technical and ethical difficulties that arise from statistical machine translation, and hope that it arouses discussions about the role of AI engineers on minimizing the harmful effects of the pressing modern concern of machine bias.

\clearpage

%\bibliographystyle{di}
\bibliography{main}

>>>>

\documentclass{article}
%\usepackage{jair, theapa, rawfonts}

%\usepackage[utf8]{inputenc}
%\usepackage{fontspec}
\usepackage{xeCJK}
%\usepackage{armtex}

@@

%\title{Is Google Translate Sexist? -- Assessing Gender Bias in Machine Translation}
%\author{\name Marcelo Prates \email morprates@inf.ufrgs.br \\
%\name Luis Lamb \email lamb@inf.ufrgs.br \\
%\addr Institute of Informatics \\
%Federal University of Rio Grande do Sul \\
%Porto Alegre, 91501-970 Brazil}

\title{Assessing Gender Bias in Machine Translation -- A Case Study with Google Translate}
\author{Marcelo Prates \and Luis Lamb}

@@

\begin{abstract}
Recently, there has been a growing concern in academia, industrial research labs and the mainstream commercial media about the phenomenon of machine bias, where trained statistical models unbeknownst to their creators become vessels of bias and prejudices. A large number of Artificial Intelligence tools have recently been suggested to be harmfully biased towards some minority, with reports of racist criminal behavior predictors, Apple's Iphone X failing to differentiate between two Asian people and the infamous case of Google photos' classifying black people as gorillas. Although a systematic study of such biases can be difficult, we believe that automated translation tools can be exploited through gender neutral languages to yield a window into the phenomenon of \emph{gender} bias in AI. 
In this paper, we construct sentences with gender neutral pronouns in different languages 
and translate these sentences into English using the Google Translate API. 
The resulting English sentences are constructed either with male, female or neutral pronouns, which allows us to collect statistics about the frequency of male defaults and evaluate which factors can influence the gender Google Translate 
API computes for each sentence. We show that not only there is a tendency towards translating pronouns to
masculine, but also that this tendency is exaggerated for sentences about male dominated fields such as Computer Science. On the other hand, sentences about artistic fields are less likely to yield male defaults. This possibly reflects the implicit assumptions our society has about stereotypical gender roles, which suggests that without proper care statistical machine translation can become hostage to the prejudices of the people behind their training datasets -- ourselves.
\end{abstract}

@@

Although the idea of automated translation can in principle be traced back to as long as the 17th century with Ren\'{e} Descartes proposal of an \say{universal language} \cite{dascal1982universal}, machine translation has only existed as a technological field since the 1950s, with a pioneering memorandum by  Warren Weaver \cite{locke1955machine,weaver1955translation} discussing the possibility of employing digital computers to perform automated translation. The now famous Georgetown-IBM experiment followed not long after, providing the first experimental demonstration of the prospects of automating translation by the means of successfully converting more than sixty Russian sentences into English \cite{gordin2015scientific}. Early systems improved upon the results of the Georgetown-IBM experiment by exploiting Noam Chomsky's theory of generative linguistics, and the field experienced a sense of optimism about the prospects of fully automating natural language translation. As is customary with artificial intelligence, the initial optimistic stage was followed by an extended period of strong disillusionment with the field, of which the catalyst was the influential 1966 ALPAC (Automatic Language Processing Advisory Committee) report( \cite{hutchins1986machine}. 
Such research was then disfavoured in the United States, making a re-entrance in the 1970s before the 1980s surge in statistical methods for machine translation \cite{koehn2009statistical,Moses2007}. Statistical and example-based machine translation have been on the rise ever since \cite{Bahdanau2014,carl2003recent,Firat2017}, with highly successful applications such as Google Translate (recently ported to a neural translation technology \cite{wu2016google}) amounting to over $200$ million users daily.

In spite of the recent commercial success of automated translation tools (or perhaps stemming directly from it), machine translation has amounted a significant deal of criticism. Noted philosopher and founding father of generative linguistics Noam Chomsky has argued that the achievements of machine translation, while successes in a particular sense, are \emph{not successes in the sense that science has ever been interested in}: they merely provide effective ways, according to Chomsky, of approximating unanalyzed data \cite{Chomsky2011,norvig2017chomsky}. Chomsky argues that the faith of the MT community in statistical methods is absurd by analogy with a standard scientific field such as physics \cite{Chomsky2011}:

\begin{quotation}
\textsl{I mean actually you could do physics this way, instead of studying things like balls rolling down frictionless planes, which can't happen in nature, if you took a ton of video tapes of what's happening outside my office window, let's say, you know, leaves flying and various things, and you did an extensive analysis of them, you would get some kind of prediction of what's likely to happen next, certainly way better than anybody in the physics department could do. Well that's a notion of success which is I think novel, I don't know of anything like it in the history of science}.
\end{quotation}

Leading AI researcher and Google's Director of Research Peter Norvig responds to these arguments by suggesting that even standard physical theories such as the Newtonian model of gravitation are, in a sense, \emph{trained} \cite{norvig2017chomsky}:

\begin{quotation}
\textsl{As another example, consider the Newtonian model of gravitational attraction, which says that the force between two objects of mass $m_1$ and $m_2$ a distance $r$ apart is given by}
\begin{equation*}
F = G m_1 m_2 / r^2
\end{equation*}
\textsl{where $G$ is the universal gravitational constant. This is a trained model because the gravitational constant G is determined by statistical inference over the results of a series of experiments that contain stochastic experimental error. It is also a deterministic (non-probabilistic) model because it states an exact functional relationship. I believe that Chomsky has no objection to this kind of statistical model. Rather, he seems to reserve his criticism for statistical models like Shannon's that have quadrillions of parameters, not just one or two.}
\end{quotation}

Chomsky and Norvig's debate \cite{norvig2017chomsky} is a microcosmos of the two leading standpoints about the future of science in the face of increasingly sophisticated statistical models. Are we, as Chomsky seems to argue, jeopardizing science by relying on statistical tools to perform predictions instead of perfecting traditional science models, or are these tools, as Norvig argues, components of the scientific standard since its conception? Currently there are no satisfactory resolutions to this conundrum, but perhaps statistical models pose an even greater and more urgent threat to our society.

On a 2014 article, Londa Schiebinger suggested that scientific research fails to take gender issues into account, arguing that the phenomenon of male defaults on new technologies such as Google Translate provides a window into this asymmetry \cite{schiebinger2014scientific}. Since then, recent worrisome results in machine learning have somewhat supported Schiebinger's view, and perhaps even partially confirmed some of Chomsky's fears. Not only Google photos' statistical image labeling algorithm has been found to classify dark-skinned people as gorillas \cite{garcia2016racist} and purportedly intelligent programs have been suggested to be negatively biased against black prisoners when predicting criminal behavior \cite{angwin2016machine} but the machine learning revolution has also indirectly revived heated debates about the controversial field of physiognomy, with proposals of AI systems capable of identifying the sexual orientation of an individual through its facial characteristics \cite{wang2017deep}. Similar concerns are growing at an unprecedented rate in the media, with reports of Apple's Iphone X face unlock feature failing to differentiate between two different Asian people \cite{womanunlockphone2017} and automatic soap dispensers which reportedly do not recognize black hands \cite{racistsoapdispenser2017}. \emph{Machine bias}, the phenomenon by which trained statistical models unbeknownst to their creators become vessels of their own prejudices, is growing into a pressing concern for the modern times, and is an invitation for us to ask ourselves whether there are limits to our dependence on these techniques -- and more importantly, whether some of these limits have already been traversed.

With this in mind, we propose a quantitative analysis of the phenomenon of gender bias in machine translation. We illustrate how this can be done by simply exploiting Google Translate to map sentences from a gender neutral language into English. As Figure \ref{fig:screenshot-gtranslate-hungarian} exemplifies, this approach produces results consistent with the hypothesis that senteces about stereotypical gender roles are translated accordingly with high probability: \emph{nurse} and \emph{baker} are translated with female pronouns while \emph{engineer} and \emph{CEO} are translated with male ones.

@@

If conclusive, in our view such an analysis would suggest that automatic translation tools should be designed in such a way as to avoid becoming hostages to their own underlying statistical models. Because these tools are trained with real world data they implicitly absorb the biases and stereotypes of our society, which must then be painstakingly removed. This poses 
mostly an ethical rather than technical limitation to statistical machine translation and big data statistical models in general.

\section{Materials and Methods}

We shall assume and then show that the phenomenon of gender bias in machine translation can be assessed by mapping sentences constructed in gender neutral languages to English by the means of an automated translation tool. Specifically, we can translate sentences such as the Hungarian \say{ő ápolónő}, where \say{ápolónő} translates to \say{nurse} and \say{ő} is a gender-neutral pronoun meaning either he, she or it, to English, yielding in this example the result \say{she's a nurse} on Google Translate. The same basic template can be ported to all other gender neutral languages, as Table \ref{tab:templates} shows. Given the success of Google Translate, which amounts to 200 million users daily, we have chosen to exploit its API to obtain the desired thermometer of gender bias. Also, in order to solidify our results, we have decided to work with as many gender neutral languages as possible, obtaining a list of these from the Wikipedia article on the issue (\url{https://en.wikipedia.org/wiki/Gender_neutrality_in_genderless_languages}). Table \ref{tab:gender-neutral-languages} compiles all languages from said article, with additional columns informing whether they
( 1) exhibit a pronominal gender system and (2) are supported by Google Translate. Because pronominal gender systems defy the purposes of our technique, such languages have been discarded.

@@

\begin{table}[H]
	\centering
	\begin{tabular}{|c|m{2cm}|m{2cm}|c|c|}
	\hline
	Language Family & Language & Pronominal Gender System & Supported & Tested \\ \hline \hline
	\multirow{2}{*}{Austronesian} 	& Malay 				& \xmark 			& \checkmark	& \checkmark	\\
									& Tagalog 				& \xmark 			& \xmark		& \xmark		\\ \hline
	\multirow{3}{*}{Finno-Ugric} 	& Estonian 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Finnish 				& \xmark 			& \checkmark	& \checkmark	\\
									& Hungarian 			& \xmark 			& \checkmark 	& \checkmark	\\ \hline
	\multirow{4}{*}{Indo-European} 	& Armenian 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Bengali 				& \xmark 			& \checkmark 	& \xmark		\\
									& English 				& \checkmark 		& \checkmark 	& \xmark 		\\
									& Persian 				& \checkmark 		& \checkmark 	& \xmark 		\\ \hline
	\multirow{3}{*}{Indo-Aryan} 	& Maithili 				& \xmark 			& \xmark 		& \xmark 		\\
									& Nepali 				& \xmark 			& \checkmark	& \xmark		\\
									& Oriya 				& \xmark 			& \xmark 		& \xmark 		\\ \hline
	\multirow{10}{*}{} 				& Japanese 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Korean 				& O 				& \checkmark 	& \xmark		\\
									& Turkish 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Yoruba 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Basque 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Swahili 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Chinese 				& O 				& \checkmark 	& \checkmark	\\
									& Cantonese 			& \xmark 			& \xmark		& \xmark		\\
									& Pipil 				& \xmark 			& \xmark		& \xmark		\\
									& Quechuan 				& \xmark 			& \xmark		& \xmark		\\ \hline
	\multirow{4}{*}{Constructed} 	& Esperanto 			& \checkmark 		& \checkmark 	& \xmark		\\
									& Ido 					& O 				& \xmark 		& \xmark 		\\
									\cline{2-2}
									& Lingua Franca Nova 	& \xmark 			& \xmark 		& \xmark 		\\
									\cline{2-2}
									& Interlingua 			& \xmark 			& \xmark 		& \xmark 		\\ \hline
	\end{tabular}
	\caption{Selected gender neutral languages obtained from the Wikipedia article \url{https://en.wikipedia.org/wiki/Gender_neutrality_in_genderless_languages}. Languages are grouped according to language families and classified according to whether they exhibit pronominal gender system (\checkmark: yes, \xmark:~no, O: it is optional). For the purposes of this work, we have decided to work only with languages lacking such a system and, as such, Persian and Esperanto have been discarded. Languages lacking support from Google Translate have been discarded.}
	\label{tab:gender-neutral-languages}
\end{table}

@@

Figures \ref{fig:histogram-female}, \ref{fig:histogram-male} and \ref{fig:histogram-neutral} shed further light on the asymmetry between the distribution male and female pronouns. While the number of occurrences of male pronouns tends towards a normal distribution, the figure changes drastically for the opposite gender, whose pronoun distribution is apparently governed by an inverse correlation. The behavior repeats itself for the gender neutral pronouns in Figure \ref{fig:histogram-neutral}.

@@

\section{Conclusions}
In this paper, we have provided evidence that statistical translation tools such as Google Translate exhibit implicit gender biases and a tendency towards male defaults, which possibly stem from the real world data used to train them. As a result, we suggest that such tools can be probed to yield insights about stereotypical gender roles in our society. By translating sentences from gender neutral languages such as Hungarian and Chinese into English, we were able to collect statistics about the asymmetry between female and male pronominal genders in the translation process. Because Google Translate typically uses English as a \emph{lingua franca} to translate between other languages, our findings probably generalize to most translations from gender neutral idioms, although we have not tested this hypothesis yet. Our results further show that although male pronouns are typical, the proportion of female pronouns can vary significantly according either to the adjectives used (\say{shy}/\say{happy} vs \say{brave}/\say{guilty}) or the category of professional occupations (artistic jobs are far more likely to be translated with female pronouns than computer-related ones). We have also shown that different languages are differently biased towards male defaults, with Hungarian exhibiting a better balance between male and female pronouns than, say, Chinese. Some languages such as Yoruba and Basque have been found to translate sentences with gender neutral pronouns very often, although this is the exception rather than the rule.

We think this work can shed further light on some of the technical and ethical difficulties that arise from statistical machine translation, and hope that it will lead to discussions about the role of AI engineers on minimizing potentially harmful effects of the pressing modern concerns of machine bias.

\bibliography{main}
\bibliographystyle{named}
